---
title: "Lab 7 Making maps"
author: "Chandana Gadela"
date: "2024-10-12"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Load required libraries
library(tidyverse)      # A collection of R packages for data manipulation and visualization
library(maps)           # Provides functions for maps and geographical data
library(mapdata)       # Extends the 'maps' package with additional data
library(readxl)        # Allows reading Excel files

```
loads the libraries that will be used for working with data and creating maps. Each library serves a specific purpose: manipulating data, working with maps, and reading Excel files.


```{r}

# Load the datasets from the specified Excel file
territories_data <- read_excel("/Users/jessy/Desktop/Info Visualization/wk7/Sales Reps.xlsx", sheet = 1)
sales_reps_data <- read_excel("/Users/jessy/Desktop/Info Visualization/wk7/Sales Reps.xlsx", sheet = 2, col_names = FALSE)
```
reads data from an Excel file. The first sheet contains information about the U.S. states and territories, while the second sheet contains information about sales representatives.


```{r}

# Clean territories data
territories <- territories_data %>%
  select(`U.S. States and Territories`, Abbreviations, `Sales Rep`) %>%  # Select relevant columns
  mutate(
    state = tolower(`U.S. States and Territories`),  # Convert state names to lowercase for consistency
    abbrev = Abbreviations,                           # Create a column for state abbreviations
    rep_id = as.character(`Sales Rep`)                # Convert Sales Rep ID to character for consistent merging
  )

```
clean up the data for U.S. states and territories. select the relevant columns (state name, abbreviation, and sales rep ID), convert the state names to lowercase for consistency, and ensure the sales rep ID is treated as text.


```{r}
# Add Canadian provinces and their sales representatives
canadian_provinces <- tribble(                       # Create a tibble for Canadian provinces
  ~state, ~abbrev, ~rep_id,
  "alberta", "AB", "7",
  "british columbia", "BC", "9",
  "manitoba", "MB", "7",
  "new brunswick", "NB", "3",
  "newfoundland and labrador", "NL", "3",
  "northwest territories", "NT", "7",
  "nova scotia", "NS", "3",
  "nunavut", "NU", NA_character_,  # No representative assigned
  "ontario", "ON", "3",
  "prince edward island", "PE", "3",
  "quebec", "QC", "3",
  "saskatchewan", "SK", "7",
  "yukon", "YT", "7"
)
```
This part manually adds data for Canadian provinces and their corresponding sales representatives. Each row includes the province name, its abbreviation, and the sales rep ID.


```{r}
# Combine US territories with Canadian provinces
territories <- bind_rows(territories, canadian_provinces)  # Bind the two datasets together

```
we combine the U.S. state data with the Canadian province data into one dataset, so we can later display both on the same map.


```{r}
# Clean sales representatives data
sales_reps <- sales_reps_data %>%
  rename(                     # Rename columns for easier access
    rep_id = ...1,           # First column as rep_id
    rep_name = ...2          # Second column as rep_name
  ) %>%
  mutate(
    rep_id = as.character(rep_id)  # Ensure rep_id is character for consistency
  )

```
chunk cleans up the sales representatives’ data. It renames the columns to make them easier to work with and ensures the sales rep IDs are treated as text.


```{r}

# Define a color palette for each representative
rep_colors <- c(
  "1" = "#E41A1C",  # Red
  "2" = "#377EB8",  # Blue
  "3" = "#4DAF4A",  # Green
  "4" = "#984EA3",  # Purple
  "5" = "#FF7F00",  # Orange
  "6" = "#FFFF33",  # Yellow
  "7" = "#A65628",  # Brown
  "8" = "#F781BF",  # Pink
  "9" = "#999999"   # Gray
)

# Get map data for US states
us_states <- map_data("state")  # Retrieves data for US states

# Get map data for Canadian provinces and territories
canada_map <- map_data("world") %>%
  filter(region %in% c("Canada", "alberta", "british columbia", "manitoba", 
                       "new brunswick", "newfoundland and labrador", 
                       "northwest territories", "nova scotia", "nunavut", 
                       "ontario", "prince edward island", "quebec", 
                       "saskatchewan", "yukon"))  # Filters for Canadian provinces

# Combine both maps (US and Canada)
us_canada_map <- bind_rows(
  us_states %>% mutate(country = "USA"),           # Label US states
  canada_map %>% mutate(country = "Canada")        # Label Canadian provinces
)

# Add rep_id and abbreviation to US states data
us_canada_map <- us_canada_map %>%
  left_join(                                         # Join the territories data to map data
    territories %>%
      select(state, rep_id, abbrev) %>%             # Select relevant columns for joining
      mutate(state = tolower(state)),                # Convert state names to lowercase
    by = c("region" = "state")                      # Match based on region and state
  )

# Calculate centers for labeling states and provinces
region_centers <- us_canada_map %>%
  group_by(region) %>%                              # Group by region
  summarize(
    long = mean(range(long)),                        # Calculate average longitude for positioning
    lat = mean(range(lat)),                          # Calculate average latitude for positioning
    rep_id = first(rep_id),                         # Take the first representative ID in the group
    abbrev = first(abbrev)                          # Take the first abbreviation in the group
  )



```


```{r}

# Create the main map with both US and Canadian regions
main_map <- ggplot() +
  # Plot regions with geom_polygon
  geom_polygon(data = us_canada_map, 
               aes(x = long, y = lat, group = group, fill = factor(rep_id)),  # Map aesthetics
               color = "white", size = 0.5) +  # White border for clarity
  # Add labels for states and provinces
  geom_text(data = region_centers,
            aes(x = long, y = lat, label = abbrev),  # Position labels at calculated centers
            size = 3.5, fontface = "bold", color = "white") +  # Formatting for labels
  # Use the custom color palette for filling regions
  scale_fill_manual(
    values = rep_colors,                           # Assign colors to reps
    name = "Sales Representatives",                 # Legend title
    labels = paste(
      sales_reps$rep_id,
      "-", sales_reps$rep_name                    # Create labels for sales reps
    )
  ) +
  # Set coordinate system for better aspect ratio
  coord_fixed(ratio = 1.3) +                      # Adjust ratio to improve map appearance
  # Customize the appearance of the map
  theme_minimal() +                                # Minimal theme for a clean look
  theme(
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),  # Title formatting
    plot.subtitle = element_text(size = 14, hjust = 0.5),                # Subtitle formatting
    legend.position = "right",                     # Position the legend
    legend.title = element_text(size = 12, face = "bold"),  # Legend title formatting
    legend.text = element_text(size = 10),        # Legend text formatting
    panel.grid = element_blank(),                  # Remove grid lines
    axis.text = element_blank(),                   # Remove axis text
    axis.title = element_blank()                   # Remove axis titles
  ) +
  # Add titles and captions to the plot
  labs(
    title = "Sales Representative Territories in US and Canada",  # Main title
    subtitle = "Each region is assigned to a distinct sales representative",  # Subtitle
    caption = paste0(                                       # Caption with additional information
      "Note: Kentucky is divided between Rep 1 and Rep 2\n",  # Note about Kentucky
      "Updated: ", format(Sys.Date(), "%B %d, %Y")        # Date of the update
    )
  )

```
map that shows both U.S. states and Canadian provinces along with their assigned sales representatives, you would merge the cleaned data with map data, assign colors to different sales reps, and label the states and provinces on the map. The rest of the code uses the combined data to generate a clear, labeled map showing the territories.


```{r}

# Print the main map to the console
print(main_map)

# Save the combined map as a PNG file with specified dimensions and resolution
ggsave("combined_us_canada_sales_map.png", main_map, width = 30, height = 15, dpi = 400)  # Increased dimensions for better clarity

```

